---
title: "Colombia COVID-19 - Central region"
author: "Angela Carraro, Giullia Monteiro Milano Oliveira, Gaia Saveri"
date: "28/07/2020"
output:
  rmdformats::readthedown:
  html_document:
    highlight: kate
    lightbox: true
    gallery: true
    toc: yes
    toc_depth: 3
  include: null
  ioslides_presentation:
    # “default”, “cerulean”, “journal”, “flatly”, “darkly”, “readable”, “spacelab”, “united”, “cosmo”, “lumen”, “paper”, “sandstone”, “simplex”, “yeti”
    highlight: kate
    theme: "sandstone"
    colortheme: "sandstone"
    lightbox: true
    widescreen: true
    smaller: true
    logo: Logo_units_blu.png #logo_dssc_alt.png
  pdf_document:
    highlight: kate
    keep_tex: yes
    toc: yes
  slidy_presentation:
    fig.height: 3
    fig.width: 4
    highlight: kate
  beamer_presentation:
    highlight: tango
    theme: "metropolis"
    colortheme: "metropolis"
    fonttheme: "structurebold"
    #always_allow_html: true
  slide_level: 2
header-includes:
- \usepackage{color}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
- \setbeamercolor{alerted text}{fg=Orange}
- \setbeamercolor{frametitle}{bg=Purple}
institute: University of Trieste
graphics: yes
fontsize: 8pt
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
```

```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
knitr::opts_chunk$set(echo = FALSE)#, tidy = TRUE)
```

```{r message=FALSE, include=FALSE}
library(MASS)
library(readr)
library(dplyr)
library(ggplot2)
library(bayesplot)
library(ggthemes)
library(ggrepel)
library(RColorBrewer)
library(leaflet)
library(geojsonio)
library(htmltools)
library(htmlwidgets)
library(rstan)
library(lubridate)
library(ciTools)
```

# Models

```{r loading, echo=TRUE}
colombia_covid <- as.data.frame(read_csv("data/covid19co.csv"))
cols <- colnames(colombia_covid)[c(1, 4, 5, 6, 7, 8, 9, 11, 14)]
colombia_covid <- colombia_covid[cols]
colombia_covid <- colombia_covid[, c(1, 9, 2, 3, 4, 5, 6, 7, 8)]
colnames(colombia_covid) <- c("ID de caso", "Fecha de diagnóstico", "Ciudad de ubicación", "Departamento o Distrito", "Atención", "Edad" , "Sexo", "Tipo", "País de procedencia")
#colombia_covid$`Departamento o Distrito`[which(colombia_covid$`Departamento o Distrito` == "Valle Del Cauca")] <- "Valle del Cauca"
central.colombia.dep <- c("Bogotá D.C.", "Tolima", "Cundinamarca", "Meta", "Boyacá", "Quindío", "Cauca",
    "Valle del Cauca", "Risaralda", "Caldas", "Boyacá", "Antioquia", "Santander", "Casanare")
central.colombia.rows <- which(colombia_covid$`Departamento o Distrito` %in% central.colombia.dep)
colombia_covid <- colombia_covid[central.colombia.rows, ]
colombia_covid <- colombia_covid[-which(colombia_covid$`Fecha de diagnóstico`== "-"), ]
head(colombia_covid, 5)
```

```{r types, include=FALSE}
## day column in "international" format (so that R can fix the sorting properly)
colombia_covid$`Fecha de diagnóstico` <- as.Date(colombia_covid$`Fecha de diagnóstico`, format="%d/%m/%Y")
colombia_covid <- colombia_covid[order(colombia_covid$`Fecha de diagnóstico`), ]
# there were missing rows since we cancelled some departments
colombia_covid$`ID de caso` <- 1:dim(colombia_covid)[1]
colombia_covid <- colombia_covid %>% mutate("Grupo de edad" = case_when(Edad <= 18 ~ '0_18',
                                                   Edad >= 19  & Edad <= 30 ~ '19_30',
                                                   Edad >=  31 & Edad <= 45 ~ '31_45',
                                                   Edad >= 46 & Edad <= 60 ~ '46_60',
                                                   Edad >=61 & Edad <= 75 ~ '60_75',
                                                   Edad >=76 ~ '76+'))
```

```{r cases}
cases <- colombia_covid %>%
  group_by(`Fecha de diagnóstico`) %>%
  count() %>% rename("Date" = `Fecha de diagnóstico`, "New cases/day" = n)
cases2 <- colombia_covid %>%
  group_by(`Fecha de diagnóstico`) %>%
  summarise(`Cumulative cases` = max(`ID de caso`))
cases <- bind_cols(cases, cases2%>%
  dplyr::select(-c(`Fecha de diagnóstico`)))
cases <- as.data.frame(cases)
cases <- cases %>% mutate(BETWEEN0 = as.numeric(difftime(Date, lag(Date, 1), units = "days")),
            BETWEEN = ifelse(is.na(BETWEEN0), 0, BETWEEN0), `Elapsed time` =
              cumsum(as.numeric(BETWEEN))) %>% dplyr::select(-c(BETWEEN0,BETWEEN))
cases <- cases[, c(1,4,2,3)]
```

```{r cases_per_dep}
cases_dep <- colombia_covid %>%
  group_by(`Fecha de diagnóstico`, `Departamento o Distrito`) %>%
  count() %>% rename("Date" = `Fecha de diagnóstico`, "Department"=`Departamento o Distrito`, "New cases/day" = n)
cases_dep <- cases_dep %>%
  group_by(`Department`) %>%
  mutate(`Cumulative cases/Department` = cumsum(`New cases/day`))
cases_dep <- as.data.frame(cases_dep)
cases_dep <- cases_dep %>% mutate(BETWEEN0 = as.numeric(difftime(Date, lag(Date, 1), units = "days")),
            BETWEEN = ifelse(is.na(BETWEEN0), 0, BETWEEN0), `Elapsed time` =
            cumsum(as.numeric(BETWEEN))) %>% dplyr::select(-c(BETWEEN0,BETWEEN))
```

```{r fix_dataset, warning=FALSE, include=FALSE}
cases_dep <- cases_dep[order(cases_dep$Department) ,]
cases_dep <- cases_dep %>%
  dplyr::mutate(dep_idx = factor(Department, levels=unique(Department)),
         `Department ID` = as.integer(dep_idx)) %>%
  dplyr::select(-dep_idx)
cases_dep <- cases_dep[, c(1, 5, 2, 6, 3, 4)]

mean_age <- colombia_covid %>%
  group_by(`Departamento o Distrito`, `Fecha de diagnóstico`) %>% 
  summarise_at(vars(Edad), funs(`Mean age` = mean(., na.rm=TRUE)))

cases_dep <- cbind(cases_dep, `Mean age` = mean_age$`Mean age`)
```

```{r dataset_Giullia, include=FALSE}
covid19 <- dplyr::select(colombia_covid, -c(`Ciudad de ubicación`,`Atención`,`Tipo`))
library(fastDummies)
covid19_dummy <- dummy_cols(covid19, select_columns = c("Departamento o Distrito", "Grupo de edad", "Sexo", "País de procedencia"), remove_first_dummy = TRUE, ignore_na=TRUE, split="-", remove_selected_columns=TRUE)
group_dummy <- covid19_dummy %>%
  group_by(`Fecha de diagnóstico`) %>%
  summarise_all(funs(sum)) %>%
  dplyr::select(-c(`Fecha de diagnóstico`,`ID de caso`))
data1 <- bind_cols(cases, group_dummy)
```

## Poisson

### Poisson with `Elapsed time` as predictor for `Cumulative cases`

```{r pois1, results = "hold", echo=TRUE, warning=FALSE}
poisson1 <- glm(`Cumulative cases` ~ `Elapsed time`,
                data=data1[1:120, ], family=poisson)
```

```{r pois1_index, results = "hold", echo=TRUE}
pred.pois1 <- poisson1$fitted.values
res.st1 <- (data1$`Cumulative cases`[1:120] - pred.pois1)/sqrt(pred.pois1)
#n=120, k=2, n-k=118
print(paste("Estimated overdispersion", sum(res.st1^2)/118))
poisson1.pred <- predict(poisson1, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson1.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson1$residuals^2))
#print(sprintf("MSE: %0.2f", sum(poisson1$residuals^2)/poisson1$df.residual))
#print(sprintf("MSE: %0.2f", anova(poisson1)['Residuals', 'Mean Sq']))
paste("AIC:", poisson1$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson1$null.deviance, deviance(poisson1)), 2))
```

### Poisson with `Elapsed time` as predictor for `New cases/day`

```{r pois1bis, results = "hold", echo=TRUE, warning=FALSE}
poisson1bis <- glm(`New cases/day` ~ `Elapsed time`,
                   data=data1[1:120, ], family=poisson)
```

```{r pois1bis_index, results = "hold", echo=TRUE}
pred.pois1bis <- poisson1bis$fitted.values
res.st1bis <- (data1$`New cases/day`[1:120] - pred.pois1bis)/sqrt(pred.pois1bis)
#n=120, k=3, n-k=117
#print(paste("Estimated overdispersion", sum(res.st1bis^2)/23))
print("DO OVERDISPERSION!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!")
poisson1bis.pred <- predict(poisson1bis, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson1bis.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson1bis$residuals^2))
paste("AIC:", poisson1bis$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson1bis$null.deviance, deviance(poisson1bis)), 2))
```

### Poisson with `Elapsed time` plus `Elapsed time`^2 as predictor

```{r pois1.5, results = "hold", echo=TRUE, warning=FALSE}
poisson1.5 <- glm(`Cumulative cases` ~ `Elapsed time` + I(`Elapsed time`^2),
                  data=cases[1:120, ], family=poisson)
```

```{r pois1.5_index, results = "hold", echo=TRUE}
pred.pois1.5 <- poisson1.5$fitted.values
res.st1.5 <- (cases$`Cumulative cases`[1:120] - pred.pois1.5)/sqrt(pred.pois1.5)
#n=120, k=3, n-k=117
print(paste("Estimated overdispersion", sum(res.st1.5^2)/117))
poisson1.5.pred <- predict(poisson1.5, newdata = cases[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson1.5.pred - cases$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson1$residuals^2))
#print(sprintf("MSE: %0.2f", sum(poisson1$residuals^2)/poisson1$df.residual))
#print(sprintf("MSE: %0.2f", anova(poisson1)['Residuals', 'Mean Sq']))
paste("AIC:", poisson1.5$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson1.5$null.deviance, deviance(poisson1.5)), 2))
```

### Poisson with `Elapsed time` plus `Sexo`

```{r pois2, results = "hold", echo=TRUE, warning=FALSE}
poisson2 <- glm(`Cumulative cases` ~ `Elapsed time` + Sexo_M,
                data=data1[1:120, ], family=poisson)
```

```{r pois2_index, results = "hold", echo=TRUE}
pred.pois2 <- poisson2$fitted.values
res.st2 <- (cases$`Cumulative cases`[1:120] - pred.pois2)/sqrt(pred.pois2)
#n=120, k=3, n-k=117
print(paste("Estimated overdispersion", sum(res.st2^2)/117))
print("DO OVERDISPERSION!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!")
poisson2.pred <- predict(poisson2, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson2.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson2$residuals^2))
paste("AIC:", poisson2$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson2$null.deviance, deviance(poisson2)), 2))
```

### Poisson with `Elapsed time` plus `Group de edad`

```{r pois3, results="hold", echo=TRUE, warning=FALSE}
poisson3 <- glm(`Cumulative cases` ~ `Elapsed time` + `Grupo de edad_19_30` +
                  `Grupo de edad_31_45` + `Grupo de edad_46_60` +
                  `Grupo de edad_60_75` + `Grupo de edad_76+`,
                data=data1[1:120, ], family=poisson)
```

```{r pois3_index, results="hold", echo=TRUE}
pred.pois3 <- poisson3$fitted.values
res.st3 <- (data1$`Cumulative cases` - pred.pois3)/sqrt(pred.pois3)
#n=120, k=7, n-k=113
print(paste("Estimated overdispersion", est.overdispersion <- sum(res.st3^2)/113))
poisson3.pred <- predict(poisson3, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson3.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson3$residuals^2))
paste("AIC:", poisson3$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson3$null.deviance, deviance(poisson3)), 2))
```

### Poisson with `Elapsed time` plus `Department`

```{r pois4, results = "hold", echo=TRUE, warning=FALSE}
poisson4 <- glm(`Cumulative cases` ~ `Elapsed time` + `Departamento o Distrito_Bogotá D.C.` +
                  `Departamento o Distrito_Boyacá` + `Departamento o Distrito_Caldas` +
                  `Departamento o Distrito_Casanare` + `Departamento o Distrito_Cauca` +
                  `Departamento o Distrito_Cundinamarca` +`Departamento o Distrito_Meta` +
                  `Departamento o Distrito_Quindío` + `Departamento o Distrito_Risaralda` +
                  `Departamento o Distrito_Santander` + `Departamento o Distrito_Tolima`,
                data=data1[1:120, ], family=poisson)
```

```{r pois4_index, results = "hold", echo=TRUE}
pred.pois4 <- poisson4$fitted.values
res.st4 <- (data1$`Cumulative cases` - pred.pois4)/sqrt(pred.pois4)
#n=120, k=7, n-k=113
print(paste("Estimated overdispersion", est.overdispersion <- sum(res.st4^2)/113))
poisson4.pred <- predict(poisson4, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson4.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson4$residuals^2))
paste("AIC:", poisson4$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson4$null.deviance, deviance(poisson4)), 2))
```

### Poisson with `Elapsed time`, `Age` and `Departments` as predictors

```{r pois5, results = "hold", echo=TRUE, warning=FALSE}
poisson5 <- glm(`Cumulative cases` ~ `Elapsed time` + `Grupo de edad_19_30` + `Grupo de edad_31_45` +
                  `Grupo de edad_46_60` + `Grupo de edad_60_75` +`Grupo de edad_76+` +
                  `Departamento o Distrito_Bogotá D.C.` + `Departamento o Distrito_Boyacá` +
                  `Departamento o Distrito_Caldas` + `Departamento o Distrito_Casanare` +
                  `Departamento o Distrito_Cauca` + `Departamento o Distrito_Cundinamarca` +
                  `Departamento o Distrito_Meta` + `Departamento o Distrito_Quindío` +
                  `Departamento o Distrito_Risaralda` + `Departamento o Distrito_Santander` +
                  `Departamento o Distrito_Tolima`, data=data1[1:120, ], family=poisson)
```

```{r pois5_index, results = "hold", echo=TRUE}
pred.pois5 <- poisson5$fitted.values
res.st5 <- (data1$`Cumulative cases` - pred.pois5)/sqrt(pred.pois5)
#n=120, k=17, n-k=103
print(paste("Estimated overdispersion", est.overdispersion <- sum(res.st5^2)/103))
poisson5.pred <- predict(poisson5, newdata = data1[120:126, ], type="response")
#paste("Real: ", data1$`Cumulative cases`[120:126], "Predict: ", poisson5.pred)
paste("RMSE:", sqrt(mean((poisson5.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson5$residuals^2))
paste("AIC:", poisson5$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson5$null.deviance, deviance(poisson5)), 2))
```

### Poisson with `Elapsed time`, `Age` and `Departments` as predictors for `New cases/day`

```{r pois5bis, results = "hold", echo=TRUE, warning=FALSE}
poisson5bis <- glm(`New cases/day` ~ `Elapsed time` + `Grupo de edad_19_30` +
                     `Grupo de edad_31_45` + `Grupo de edad_46_60` + `Grupo de edad_60_75` +
                     `Grupo de edad_76+` + `Departamento o Distrito_Bogotá D.C.` +
                     `Departamento o Distrito_Boyacá` + `Departamento o Distrito_Caldas` +
                     `Departamento o Distrito_Casanare` + `Departamento o Distrito_Cauca` +
                     `Departamento o Distrito_Cundinamarca` + `Departamento o Distrito_Meta` +
                     `Departamento o Distrito_Quindío` + `Departamento o Distrito_Risaralda` +
                     `Departamento o Distrito_Santander` + `Departamento o Distrito_Tolima`,
                   data=data1[1:120, ], family=poisson)
```

```{r pois5bis_index, results = "hold", echo=TRUE}
pred.pois5bis <- poisson5bis$fitted.values
res.st5bis <- (data1$`Cumulative cases` - pred.pois5bis)/sqrt(pred.pois5bis)
#n=120, k=18, n-k=102
print(paste("Estimated overdispersion", est.overdispersion <- sum(res.st5bis^2)/102))
poisson5bis.pred <- predict(poisson5bis, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson5bis.pred - data1$`New cases/day`[120:126])^2)))
#paste("MSE:", mean(poisson5$residuals^2))
paste("AIC:", poisson5bis$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson5bis$null.deviance, deviance(poisson5bis)), 2))
```

### Poisson with `Elapsed time`, `Elapsed time`^2, `Age` and `Departments` as predictors

```{r pois6, results = "hold", echo=TRUE, warning=FALSE}
poisson6 <- glm(`Cumulative cases` ~ `Elapsed time` + I(`Elapsed time`^2) +
                  `Grupo de edad_19_30` + `Grupo de edad_31_45` + `Grupo de edad_46_60` +
                  `Grupo de edad_60_75` + `Grupo de edad_76+` + `Departamento o Distrito_Bogotá D.C.` +
                  `Departamento o Distrito_Boyacá` + `Departamento o Distrito_Caldas` +
                  `Departamento o Distrito_Casanare` + `Departamento o Distrito_Cauca` +
                  `Departamento o Distrito_Cundinamarca` + `Departamento o Distrito_Meta` +
                  `Departamento o Distrito_Quindío` + `Departamento o Distrito_Risaralda` +
                  `Departamento o Distrito_Santander` + `Departamento o Distrito_Tolima`,
                data=data1[1:120, ], family=poisson)
```

```{r pois6_index, results = "hold", echo=TRUE}
pred.pois6 <- poisson6$fitted.values
res.st6 <- (data1$`Cumulative cases` - pred.pois6)/sqrt(pred.pois6)
#n=120, k=19, n-k=101
print(paste("Estimated overdispersion", est.overdispersion <- sum(res.st6^2)/101))
poisson6.pred <- predict(poisson6, newdata = data1[120:126, ], type="response")
#paste("Real: ", data1$`Cumulative cases`[120:126], "Predict: ", poisson6.pred)
paste("RMSE:", sqrt(mean((poisson6.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson4$residuals^2))
paste("AIC:", poisson6$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson6$null.deviance, deviance(poisson6)), 2))
```

### Summary table - Poisson models for `Cumulative cases`

```{r summary_poisson_cumulative, echo=FALSE, warning=FALSE}
library(kableExtra)
poisson.cumulative.summary<-data.frame(Model="Cumulative cases ~ Elapsed time", AIC = 21915, RMSE = 1760)
poisson1.5.summary<-c("Cumulative cases ~ Elapsed time + Elapsed time^2", 12062, 6940)
poisson2.summary<-c("Cumulative cases ~ Elapsed time + Sex", 20887, 3852)
poisson3.summary<-c("Cumulative cases ~ Elapsed time + Age group", 20650, 3249)
poisson4.summary<-c("Cumulative cases ~ Elapsed time + Department", 19275, 3328)
poisson5.summary<-c("Cumulative cases ~ Elapsed time + Department + Age group", 18085, 8243)
poisson6.summary<-c("Cumulative cases ~ Elapsed time + Elapsed time^2 + Department + Age group", 8781, 8441)

poisson.cumulative.summary$Model<-as.character(poisson.cumulative.summary$Model)
poisson.cumulative.summary<-rbind(poisson.cumulative.summary, poisson1.5.summary)
poisson.cumulative.summary<-rbind(poisson.cumulative.summary, poisson2.summary)
poisson.cumulative.summary<-rbind(poisson.cumulative.summary, poisson3.summary)
poisson.cumulative.summary<-rbind(poisson.cumulative.summary, poisson4.summary)
poisson.cumulative.summary<-rbind(poisson.cumulative.summary, poisson5.summary)
poisson.cumulative.summary<-rbind(poisson.cumulative.summary, poisson6.summary)

kable(poisson.cumulative.summary) %>%
  kable_styling(bootstrap_options = "striped")
```

### Summary table - Poisson models for `New cases/day`

```{r summary_poisson_cases, echo=FALSE, warning=FALSE}
poisson.cases.summary<-data.frame(Model="New cases/day ~ Elapsed time", AIC = 8427, RMSE = 58642)
poisson5bis.summary<-c("New cases/day ~ Elapsed time + Age group + Department", 2860, 1225)

poisson.cases.summary$Model<-as.character(poisson.cases.summary$Model)
poisson.cases.summary<-rbind(poisson.cases.summary, poisson5bis.summary)

kable(poisson.cases.summary) %>%
  kable_styling(bootstrap_options = "striped")
```

### ANOVA to compare the Poisson models

```{r Angela11, echo=TRUE, warning=FALSE}
anova(poisson1, poisson2, poisson3, poisson4, poisson5, test="Chisq")
```


## Quasi-Poisson

### Quasi Poisson with `Elapsed time` as predictor

```{r quasi_pois, results = "hold", echo=TRUE, warning=FALSE}
poisson1quasi <- glm(`Cumulative cases` ~ `Elapsed time`,
                     data=data1[1:120, ], family=quasipoisson)
```

```{r quasi_pois_index, results = "hold", echo=TRUE}
pred.poisq <- poisson1quasi$fitted.values
res.stq <- (data1$`Cumulative cases` - pred.poisq)/sqrt(summary(poisson1quasi)$dispersion*pred.poisq)
#n=120, k= ?, n-k=?
print(paste("Estimated overdispersion", sum(res.stq^2)/23))
poisson1quasi.pred <- predict(poisson1quasi, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((poisson1quasi.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson1quasi$residuals^2))
paste("AIC:", poisson1quasi$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson1quasi$null.deviance, deviance(poisson1quasi)), 2))
```

### Quasi Poisson with `Elapsed time` as predictors for `New cases/day`

```{r pois1qbis, results = "hold", echo=TRUE, warning=FALSE}
poisson1quasibis <- glm(`New cases/day` ~ `Elapsed time`,
                        data=data1[1:120, ], family=quasipoisson)
```

```{r pois1qbis_index, results = "hold", echo=TRUE}
pred.pois1quasibis <- poisson1quasibis$fitted.values
res.st1quasibis <- (data1$`Cumulative cases` - pred.pois1quasibis)/sqrt(pred.pois1quasibis)
#n=120, k=18, n-k=102
print(paste("Estimated overdispersion", est.overdispersion <- sum(res.st1quasibis^2)/102))
poisson1quasibis.pred <- predict(poisson1quasibis, newdata = data1[120:126, ], type="response")
paste("RMSE:", sqrt(mean((poisson1quasibis.pred - data1$`New cases/day`[120:126])^2)))
#paste("MSE:", mean(poisson1quasibis$residuals^2))
paste("AIC:", poisson1quasibis$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson1quasibis$null.deviance, deviance(poisson1quasibis)), 2))
```

### Quasi Poisson with `Elapsed time` and `Age` as predictor

```{r quasi_pois2, results = "hold", echo=TRUE, warning=FALSE}
poisson2quasi <- glm(`Cumulative cases` ~ `Elapsed time` + `Grupo de edad_19_30` +
                       `Grupo de edad_31_45` + `Grupo de edad_46_60` +
                       `Grupo de edad_60_75` + `Grupo de edad_76+`,
                     data=data1[1:120, ], family=quasipoisson)
```

```{r quasi_pois2_index, results = "hold", echo=TRUE}
pred.poisq2 <- poisson2quasi$fitted.values
res.stq2 <- (data1$`Cumulative cases` - pred.poisq2)/sqrt(summary(poisson2quasi)$dispersion*pred.poisq2)
#n=120, k= ?, n-k=?
print(paste("Estimated overdispersion", sum(res.stq2^2)/18))
poisson2quasi.pred <- predict(poisson2quasi, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((poisson2quasi.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(poisson2quasi$residuals^2))
paste("AIC:", poisson2quasi$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(poisson2quasi$null.deviance, deviance(poisson2quasi)), 2))
```

### Summary table - Quasi Poisson models for `Cumulative cases`

```{r summary_quasi_poisson_cumulative, echo=FALSE, warning=FALSE}
quasipoisson.cumulative.summary<-data.frame(Model="Cumulative cases ~ Elapsed time", RMSE = 1760)
poisson2quasi.summary<-c("Cumulative cases ~ Elapsed time + Age group", 3249)

quasipoisson.cumulative.summary$Model<-as.character(quasipoisson.cumulative.summary$Model)
quasipoisson.cumulative.summary<-rbind(quasipoisson.cumulative.summary, poisson2quasi.summary)

kable(quasipoisson.cumulative.summary) %>%
  kable_styling(bootstrap_options = "striped")
```

### Summary table - Quasi Poisson models for `New cases/day`

```{r summary_quasi_poisson_cases, echo=FALSE, warning=FALSE}
quasipoisson.cases.summary<-data.frame(Model="New cases/day ~ Elapsed time", RMSE = 753)

quasipoisson.cases.summary$Model<-as.character(quasipoisson.cases.summary$Model)

kable(quasipoisson.cases.summary) %>%
  kable_styling(bootstrap_options = "striped")
```


## Negative Binomial

### Negative Binomial with `Elapsed time` as predictor

```{r nb1, results = "hold", echo=TRUE, warning=FALSE}
nb1 <- glm.nb(`Cumulative cases` ~ `Elapsed time`,
              data=data1[1:120, ])
```

```{r nb1_index, results = "hold", echo=TRUE}
#n=120, k=2, n-k=118
stdres <- rstandard(nb1)
print(paste("Estimated overdispersion", sum(stdres^2)/118))
nb1.pred <- predict(nb1, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((nb1.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(nb1$residuals^2))
paste("AIC:", nb1$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(nb1$null.deviance, deviance(nb1)), 2))
```

### Negative Binomial with `Elapsed time` plus `Age` as predictors

```{r nb2, results = "hold", echo=TRUE, warning=FALSE}
nb2 <- glm.nb(`Cumulative cases` ~ `Elapsed time` + `Grupo de edad_19_30`+
                `Grupo de edad_31_45` + `Grupo de edad_46_60` +
                `Grupo de edad_60_75` + `Grupo de edad_76+`,
              data=data1[1:120, ])
```

```{r nb2_index, results = "hold", echo=TRUE}
nb2.pred <- predict(nb2, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((nb2.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(nb2$residuals^2))
paste("AIC:", nb2$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(nb2$null.deviance, deviance(nb2)), 2))
```

### Negative Binomial with `Elapsed time` plus `Department` as predictors

```{r nb3, results = "hold", echo=TRUE, warning=FALSE}
nb3 <- glm.nb(`Cumulative cases` ~ `Elapsed time` + `Departamento o Distrito_Bogotá D.C.` +
                `Departamento o Distrito_Boyacá` + `Departamento o Distrito_Caldas`+
                `Departamento o Distrito_Casanare` + `Departamento o Distrito_Cauca` +
                `Departamento o Distrito_Cundinamarca` + `Departamento o Distrito_Meta` +
                `Departamento o Distrito_Quindío` + `Departamento o Distrito_Risaralda` +
                `Departamento o Distrito_Santander` + `Departamento o Distrito_Tolima`,
              data=data1[1:120, ])
```

```{r nb3_index, results = "hold", echo=TRUE}
nb3.pred <- predict(nb3, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((nb3.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(nb3$residuals^2))
paste("AIC:", nb3$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(nb3$null.deviance, deviance(nb3)), 2))
```

### Negative Binomial with `Elapsed time`, `Age` and `Departments` as pedictors

```{r nb4, results = "hold", echo=TRUE, warning=FALSE}
nb4 <- glm.nb(`Cumulative cases` ~ `Elapsed time` + `Grupo de edad_19_30` + `Grupo de edad_31_45` +
                `Grupo de edad_46_60` + `Grupo de edad_60_75` + `Grupo de edad_76+` +
                `Departamento o Distrito_Bogotá D.C.` + `Departamento o Distrito_Boyacá` +
                `Departamento o Distrito_Caldas` + `Departamento o Distrito_Casanare` +
                `Departamento o Distrito_Cauca` + `Departamento o Distrito_Cundinamarca` +
                `Departamento o Distrito_Meta` + `Departamento o Distrito_Quindío` +
                `Departamento o Distrito_Risaralda` + `Departamento o Distrito_Santander` +
                `Departamento o Distrito_Tolima`, data=data1[1:120, ])
```

```{r nb4_index, results = "hold", echo=TRUE}
# Calculating overdispersion n=120 k=19 n-k=101
stdres <- rstandard(nb4)
print(paste("Estimated overdispersion", sum(stdres^2)/101))
nb4.pred <- predict(nb4, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((nb4.pred - data1$`Cumulative cases`[120:126])^2)))
#paste("MSE:", mean(nb5$residuals^2))
paste("AIC:", nb4$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(nb4$null.deviance, deviance(nb4)), 2))
```


### Negative Binomial with `Elapsed time`, `Age` and `Departments` as pedictors  for `New cases/day`

```{r nb4bis, results = "hold", echo=TRUE, warning=FALSE}
nb4bis <- glm.nb(`New cases/day` ~ `Elapsed time` + `Grupo de edad_19_30` + `Grupo de edad_31_45` +
                   `Grupo de edad_46_60` + `Grupo de edad_60_75` + `Grupo de edad_76+` +
                   `Departamento o Distrito_Bogotá D.C.` + `Departamento o Distrito_Boyacá` +
                   `Departamento o Distrito_Caldas` + `Departamento o Distrito_Casanare` +
                   `Departamento o Distrito_Cauca` + `Departamento o Distrito_Cundinamarca` +
                   `Departamento o Distrito_Meta` + `Departamento o Distrito_Quindío` +
                   `Departamento o Distrito_Risaralda` + `Departamento o Distrito_Santander` +
                   `Departamento o Distrito_Tolima`, data=data1[1:120, ])
```

```{r nb4bis_index, results = "hold", echo=TRUE}
# Calculating overdispersion n=120 k=19 n-k=101
stdres2 <- rstandard(nb4bis)
print(paste("Estimated overdispersion", sum(stdres2^2)/101))
nb4bis.pred <- predict(nb4bis, newdata = data1[120:126, ], type = "response")
paste("RMSE:", sqrt(mean((nb4bis.pred - data1$`New cases/day`[120:126])^2)))
#paste("MSE:", mean(nb4bis$residuals^2))
paste("AIC:", nb4bis$aic)
paste(c("Null deviance: ", "Residual deviance:"),
       round(c(nb4bis$null.deviance, deviance(nb4bis)), 2))
```


### Summary table - Negative Binomial models for `Cumulative cases`

```{r summary_nb_cumulative, echo=FALSE, warning=FALSE}
nb.cumulative.summary<-data.frame(Model="Cumulative cases ~ Elapsed time", AIC = 21911, RMSE = 1765)
nb2.summary<-c("Cumulative cases ~ Elapsed time + Age group", 20645, 3253)
nb3.summary<-c("Cumulative cases ~ Elapsed time + Department", 19270, 3335)
nb4.summary<-c("Cumulative cases ~ Elapsed time + Department + Age group", 1808, 8253)

nb.cumulative.summary$Model<-as.character(nb.cumulative.summary$Model)
nb.cumulative.summary<-rbind(nb.cumulative.summary, nb2.summary)
nb.cumulative.summary<-rbind(nb.cumulative.summary, nb3.summary)
nb.cumulative.summary<-rbind(nb.cumulative.summary, nb4.summary)

kable(nb.cumulative.summary) %>%
  kable_styling(bootstrap_options = "striped")
```

### Summary table - Negative Binomial models for `New cases/day`

```{r summary_nb_cases, echo=FALSE, warning=FALSE}
nb.cases.summary<-data.frame(Model="New cases/day ~ Elapsed time + Age group + Department", AIC = 1435, RMSE = 1379)

nb.cases.summary$Model<-as.character(nb.cases.summary$Model)

kable(nb.cases.summary) %>%
  kable_styling(bootstrap_options = "striped")
```

### Applying ANOVA to compare the negative binomial models

```{r anova_nb, echo=TRUE, warning=FALSE}
#Applying ANOVA to compare the negative binomial models
anova(nb1, nb2, nb3, nb4)
```