---
title: "Colombia COVID-19 - Central region"
author: "Angela Carraro, Giullia Monteiro Milano Oliveira, Gaia Saveri"
date: "3/07/2020"
output:
  ioslides_presentation:
    highlight: kate
    lightbox: true
    widescreen: true
    smaller: true
    logo: Logo_units_blu.png #logo_dssc_alt.png
  rmdformats::readthedown:
  beamer_presentation:
    highlight: tango
    theme: "metropolis"
    colortheme: "metropolis"
    fonttheme: "structurebold"
    always_allow_html: true
  slidy_presentation:
    fig.height: 3
    fig.width: 4
    highlight: kate
  html_document:
    highlight: kate
    lightbox: true
    gallery: true
    toc: yes
    toc_depth: 3
  include: null
  pdf_document:
    highlight: kate
    keep_tex: yes
    toc: yes
  slide_level: 2
header-includes:
- \usepackage{color}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
- \setbeamercolor{alerted text}{fg=Orange}
- \setbeamercolor{frametitle}{bg=Purple}
institute: University of Trieste
graphics: yes
fontsize: 8pt
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
```

```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

```{r message=FALSE, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(leaflet)
library(geojsonio)
library(htmltools)
library(htmlwidgets)
```

## Our project

We decided to do **central Colombia**, basically because it is where the capital is.

We built a model for the number of `confirmed` cases using all the others covariates (plus some we created) and we estimated the predictive accuracy of our selected model.

## Our cities

We decided to consider as **central Colombia** the following cities:

 - Bogotà DC,
 - Boyacá,
 - Tolima,
 - Cundinamarca,
 - Meta,
 - Quindío,
 - Valle del Cauca,
 - Risaralda, Celdas,
 - Boyacá,
 - Antioquia,
 - Santander
 - Casanare

## Loading the dataset

```{r loading}
colombia_covid <- as.data.frame(read_csv("data/datasets_567855_1056808_Casos1.csv"))
colnames(colombia_covid)[5] <- "Atención"
colnames(colombia_covid)[8] <- "Tipo"
# slicing the main dataset
central.colombia.dep <- c("Bogotá D.C.", "Tolima", "Cundinamarca", "Meta", "Boyacá", "Quindío", "Cauca",
    "Valle del Cauca", "Risaralda", "Caldas", "Boyacá", "Antioquia", "Santander", "Casanare")
central.colombia.rows <- which(colombia_covid$`Departamento o Distrito` %in% central.colombia.dep)
colombia_covid <- colombia_covid[central.colombia.rows, ]
```

## Map

The color of the pins is related with the number of cases: if they are less than 10 the color is "green", if they are less than 100 the color is "orange", otherwise it is "red".\

```{r create_gis_df, include=FALSE}
#lat-long
#bogota<c(4.592164298, -74.072166378, 542)
valle_de_cauca<-c("Valle del Cauca", 3.4372200, -76.5225000, 150)
cauca<-c("Cauca", 2.43823, -76.6131592, 12) 
antioquia<-c("Antioquia",6.2518400, -75.5635900, 127)
cartagena<-c("Cartagena D.T. y C", 10.39972, -75.51444, 39)
huila<-c("Huila", 2.9273, -75.2818909, 30)
meta<-c("Meta", 4.1420000, -73.6266400, 12)
risaralda<-c("Risaralda", 4.8133302, -75.6961136, 35)
norte_santander<-c("Norte de Santander", 7.8939100, -72.5078200, 21)
caldas<-c("Caldas", 5.0688900, -75.5173800, 16)
cudinamarca<-c("Cundinamarca", 4.862437, -74.058655, 42)
barraquilla<-c("Barranquilla D.E.", 10.9685400, -74.7813200, 35) #atlantico
santader<-c("Santander", 7.1253900, -73.1198000, 12)
quindio<-c("Quindío", 4.535000, -75.675690, 23)
tolima<-c("Tolima", 4.43889, -75.2322235, 14)
santa_marta<-c("Santa Marta D.T. y C.", 11.24079, -74.19904, 12)
cesar<-c("Cesar", 10.4631400, -73.2532200, 16)
san_andres<-c("San Andrés", 12.5847197, -81.7005615, 2)
casanare<-c("Casanare", 5.3377500, -72.3958600, 2)
narino<-c("Nariño", 1.2136100, -77.2811100, 6)
boyaca<-c("Boyacá", 5.767222, -72.940651, 6)
cordoba<-c("Córdoba", 8.7479800, -75.8814300, 2)
bolivar<-c("Bolívar", 10.3997200, -75.5144400, 3)
sucre<-c("Sucre", 9.3047199, -75.3977814, 1)
guajira<-c("La Guajira", 11.5444400, -72.9072200, 1)

gis_data<-data.frame(name="Bogotá D.C.", latitude=4.624335, longitude=-74.063644, cases=542) #bogotà
gis_data$name<-as.character(gis_data$name)
gis_data<-rbind(gis_data, cauca)
gis_data<-rbind(gis_data, valle_de_cauca)
gis_data<-rbind(gis_data, antioquia)
gis_data<-rbind(gis_data, cartagena)
gis_data<-rbind(gis_data, huila)
gis_data<-rbind(gis_data, meta)
gis_data<-rbind(gis_data, risaralda)
gis_data<-rbind(gis_data, norte_santander)
gis_data<-rbind(gis_data, caldas)
gis_data<-rbind(gis_data, cudinamarca)
gis_data<-rbind(gis_data, barraquilla)
gis_data<-rbind(gis_data, santader)
gis_data<-rbind(gis_data, quindio)
gis_data<-rbind(gis_data, tolima)
gis_data<-rbind(gis_data, santa_marta)
gis_data<-rbind(gis_data, cesar)
gis_data<-rbind(gis_data, san_andres)
gis_data<-rbind(gis_data, casanare)
gis_data<-rbind(gis_data, narino)
gis_data<-rbind(gis_data, boyaca)
gis_data<-rbind(gis_data, cordoba)
gis_data<-rbind(gis_data, bolivar)
gis_data<-rbind(gis_data, sucre)
gis_data<-rbind(gis_data, guajira)

gis_data$latitude<-as.numeric(gis_data$latitude)
gis_data$longitude<-as.numeric(gis_data$longitude)
gis_data$cases<-as.numeric(gis_data$cases)
```

```{r central_colombia, echo=FALSE, inlcude=FALSE}
dept<-geojsonio::geojson_read("data/Colombia.geo.json", what="sp")
#slice the gis_data dataset
central_gis_data<-gis_data[which(gis_data$name %in% central.colombia.dep),]
#slice the geojson dataset 
central_dept<-c("SANTAFE DE BOGOTA D.C", "TOLIMA", "CUNDINAMARCA", "META", "BOYACA", "QUINDIO", "CAUCA", "VALLE DEL CAUCA", "RISARALDA", "CALDAS", "BOYACA", "ANTIOQUIA", "SANTANDER", "CASANARE")
central.dept<-dept[which(dept@data$NOMBRE_DPT %in% central_dept),]
```

```{r our_map, echo=TRUE, warning=FALSE, include=FALSE}
getColor <- function(central_gis_data) {
  sapply(central_gis_data$cases, function(cases) {
  if(cases <= 10) {
    "green"
  } else if(cases <= 100) {
    "orange"
  } else {
    "red"
  } })
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(central_gis_data)
)

#now colombia is yellow and our departments are red

central_map <- leaflet(data=central_gis_data) %>% addTiles() %>%
  addAwesomeMarkers(~longitude, ~latitude, icon=icons, label = ~htmlEscape(paste(name,cases,sep=" : "))) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data=dept,
              opacity=0.1,
              color="yellow",
              fillOpacity = 0.1) %>%
  addPolygons(data=central.dept,     
              opacity = 0.2,
              color = "red",
              dashArray = "3",
              fillOpacity = 0.1)
#saveWidget(central_map, file="figs/central_colombia.html")
```

```{r}
central_map
```


## Preprocessing {.smaller}

```{r types, include=FALSE}
colombia_covid$`Fecha de diagnóstico` <- as.Date(colombia_covid$`Fecha de diagnóstico`, format="%d/%m/%Y")
## recodifying distinct times
l_times <- length(unique(colombia_covid$`Fecha de diagnóstico`))
times <- as.factor(colombia_covid$`Fecha de diagnóstico`)
levels(times) <- c(1:l_times)
colombia_covid$Veces <- as.numeric(times)
# The max of the `ID de caso` for each date is the cumulative number of cases confirmed up to that date
#acumulativo <- colombia_covid %>%
#  group_by(`Fecha de diagnóstico`) %>%s
#  summarise(max(`ID de caso`))
#colnames(acumulativo)[2] <- "cumulative"
#colombia_covid$Acumulativo <- colombia_covid$`Fecha de diagnóstico` == acumulativo$`Fecha de diagnóstico`
colombia_covid <- colombia_covid %>% mutate("Grupo de edad" = case_when(Edad <= 18 ~ '0-18',
                                                   Edad >= 19  & Edad <= 30 ~ '19-30',
                                                   Edad >=  31 & Edad <= 45 ~ '31-45',
                                                   Edad >= 46 & Edad <= 60 ~ '46-60',
                                                   Edad >=61 & Edad <= 75 ~ '60-75',
                                                   Edad >=76 ~ '76+'))
colombia_covid$`Grupo de edad` <- as.factor(colombia_covid$`Grupo de edad`)
colombia_covid$`Departamento o Distrito` <- as.factor(colombia_covid$`Departamento o Distrito`)
colombia_covid$`Ciudad de ubicación` <- as.factor(colombia_covid$`Ciudad de ubicación`)
colombia_covid$Sexo <- as.factor(colombia_covid$Sexo)
colombia_covid$Atención <- as.factor(colombia_covid$Atención)
colombia_covid$Tipo <- as.factor(colombia_covid$Tipo)
```

```{r fix_countries, include=FALSE}
sort(unique(colombia_covid$`País de procedencia`))
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Isla Martin - Caribe"] <- "Islas San Martin"
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Israel Egipto"] <- "Israel - Egipto"
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Jamaica - Isla Caimán - Panamá"] <- "Jamaica - Panamá - Isla Caimán"
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Madrid"] <- "España"
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Londres"] <- "Inglaterra"
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Alemania - Estambul"] <- "Alemania - Turquía"
colombia_covid$`País de procedencia`[colombia_covid$`País de procedencia` == "Jamaica - Panamá - Islas del caribe - Cartagena"] <- "Jamaica - Panamá - Colombia"
colombia_covid$`País de procedencia` <- as.factor(colombia_covid$`País de procedencia`)
```

```{r create_continents, include=FALSE}
Europa <- c("Alemania", "Bélgica", "Europa",  "Croacia", "España", "España - Croacia", "España - Croacia - Bosnia",  "España - Francia", "España - Italia", "Francia", "Francia - Holanda", "Grecia", "Inglaterra", "Italia", "Italia - Ucrania - España", "Suiza")
Asia <- c("Arabia", "Emiratos Árabes", "Turquía")
África <- c("Egipto", "Marruecos")
Norteamérica <- c("Canadá", "Estados Unidos", "México")
Centroamérica <- c("Aruba", "Costa Rica", "Cuba", "Guatemala", "Islas San Martin", "Jamaica", "Jamaica - Panamá - Isla Caimán", "Jamaica - Panamá - Islas del caribe - Cartagena", "Panamá", "Panamá - Jamaica", "Puerto Rico", "República Dominicana")
Sudamerica <- c("Argentina", "Brasil", "Chile", "Ecuador", "Perú", "Venezuela")
# Alemania - Turquía", "España - India", "España - Turquia", "Italia - España - Turquía", "Turquía - Grecia" -> "Europa - Asia"
# "España - Egipto" -> "Europa - África"
# "Israel - Egipto" -> "Asia - África"
# "Italia - Jamaica - Panamá" -> "Europa - Centroamérica"
# "Colombia" -> "Colombia"

for (i in 1:nrow(colombia_covid)) {
  if (colombia_covid$`País de procedencia`[i] %in% Europa){
    colombia_covid$`Continente de procedencia`[i] <- "Europa"}
  else if (colombia_covid$`País de procedencia`[i] %in% Asia){
    colombia_covid$`Continente de procedencia`[i] <- "Asia"}
  else if (colombia_covid$`País de procedencia`[i] %in% África){
    colombia_covid$`Continente de procedencia`[i] <- "África"}
  else if (colombia_covid$`País de procedencia`[i] %in% Norteamérica){
    colombia_covid$`Continente de procedencia`[i] <- "Norteamérica"}
  else if (colombia_covid$`País de procedencia`[i] %in% Centroamérica){
    colombia_covid$`Continente de procedencia`[i] <- "Centroamérica"}
  else if (colombia_covid$`País de procedencia`[i] %in% Sudamerica){
    colombia_covid$`Continente de procedencia`[i] <- "Sudamerica"}
  else if (colombia_covid$`País de procedencia`[i] == "Colombia"){
    colombia_covid$`Continente de procedencia`[i] <- "Colombia"}
  else if (colombia_covid$`País de procedencia`[i] == "Alemania - Turquía"){
    colombia_covid$`Continente de procedencia`[i] <- "Europa - Asia"}
  else if (colombia_covid$`País de procedencia`[i] == "España - India")
    colombia_covid$`Continente de procedencia`[i] <- "Europa - Asia"
  else if (colombia_covid$`País de procedencia`[i] == "España - Turquia")
    colombia_covid$`Continente de procedencia`[i] <- "Europa - Asia"
  else if (colombia_covid$`País de procedencia`[i] == "Italia - España - Turquía")
    colombia_covid$`Continente de procedencia`[i] <- "Europa - Asia"
  else if (colombia_covid$`País de procedencia`[i] == "Turquía - Grecia")
    colombia_covid$`Continente de procedencia`[i] <- "Europa - Asia"
  else if (colombia_covid$`País de procedencia`[i] == "España - Egipto")
    colombia_covid$`Continente de procedencia`[i] <- "Europa - África"
  else if (colombia_covid$`País de procedencia`[i] == "Israel - Egipto")
    colombia_covid$`Continente de procedencia`[i] <- "Asia - África"
  else if (colombia_covid$`País de procedencia`[i] == "Italia - Jamaica - Panamá")
    colombia_covid$`Continente de procedencia`[i] <- "Europa - Centroamérica"
}
colombia_covid$`Continente de procedencia` <- as.factor(colombia_covid$`Continente de procedencia`)
# Transforming the 0 value into a null value
library(naniar)
colombia_covid <- colombia_covid %>% replace_with_na(replace = list(`País de procedencia` = 0))
# colombia_covid[911,]$`País de procedencia` <- NA
```

```{r display}
head(colombia_covid)
```