---
title: "SIR temptative model"
author: "Angela Carraro, Giullia Monteiro Milano Oliveira, Gaia Saveri"
date: "3/07/2020"
output:
  html_document:
    highlight: kate
    lightbox: true
    gallery: true
    toc: yes
    toc_depth: 3
  ioslides_presentation:
    highlight: kate
  rmdformats::readthedown:
  include: null
  beamer_presentation:
    highlight: kate
  pdf_document:
    highlight: kate
    keep_tex: yes
    toc: yes
  slide_level: 2
  slidy_presentation:
    fig.height: 3
    fig.width: 4
    highlight: kate
header-includes:
- \usepackage{color}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
- \setbeamercolor{alerted text}{fg=Orange}
- \setbeamercolor{frametitle}{bg=Purple}
institute: University of Udine & University of Trieste
graphics: yes
fontsize: 10pt
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
```

```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

```{r message=FALSE}
library(readr)
library(rstan)
library(dplyr)
library(ggplot2)
library(bayesplot)
library(lubridate)
```

## The SIR model 

The SIR model (which is an acronym for Susceptible, Infected, Recovered or Removed) is a *population-based* model, which subdivide the population into homogeneous time-dependent groups, called *compartments*. The flows between compartments can be described by a system of ODEs. 

The temporal dynamic of the sizes of each of the compartmnerts are governed by the following system of ODEs:

$$
\frac{dS}{dt} = -\beta\cdot S \cdot \frac{I}{N} \\
\frac{dI}{dt} = \beta \cdot S \cdot \frac{I}{N} - \gamma\cdot I \\
\frac{dR}{dt} = \gamma\cdot I
$$

where $S(t)$ is the number of people suscptible to becoming infected, $I(t)$ is the number of people currently infected, $R(t)$ is the number of people that are recovered or removed (dead). 

The $\beta$ is called *effective contact rate* and controls the rate of spread which represents the probability of transmitting disease between a susceptible and an infectious individual. 

The $\gamma$ is called *recoverity and mortality* rate and it is determined by the average duration of infection. 

Assumption behind this model are that:

  * $N=S+I+R$ remains constant, namely births and deaths (except the ones due to the disease) are not contributing to the dynamics;
  
  * recovered individual cannot be susceptible again (full immunity);
  
  * infection rate and recovery rate are constant;
  
  * individuals meet any other individual uniformely at random and recovery time follows an exponential distribution woth mean $\frac{1}{\gamma}$.
  
This is the mathematical framework of the SIR model, now we need to give it a statistical formulation. 

## Statistical model

Let's start by the likelihood, also called *sampling distribution* $p(Y|\theta)$ which tells us, given the parameters, how to generate the data. We want to link the solution we get from the ODEs to the observed data. Let's say that $\hat{I}(t)$ are the number of infected people at time $t$ as predictedd by the SIR model, while $I(t)$ is the number of observed infected people at time $t$ (comfirmed cases as given by our data). 

As we proposed in other models (the class of glm) we can model the number of infected people with a count distribution, such as the Negative Binomial, having $\hat{I}(t)$ as expected value and a parameter $\phi$ that allows for overdispersion. 

$$
I(t) \sim \mathcal{NagativeBinomial}(\hat{I}(t), \phi)
$$

With this model, the set of parameters we have to deal with now is $\theta = (\beta, \gamma, \phi)$. 

Being our model a Bayesian one, we need to put a prior (which encodes the information about the parameters we have before observing the data) over these parameters, $p(\theta)$.  

We can use non-informative priors for them such as:

$$
\beta \sim \mathcal{N}(2,1) \\
\gamma \sim \mathcal{N}(0.4, 0.5) \\
\frac{1}{\phi} \sim \mathcal{Exp}(5)
$$

Having the likelihood and the prior, we are at this point ready to apply Bayes theorem to estimate the posterior $p(\theta|Y) \propto p(Y|theta)\cdot p(\theta)$, by fitting the model.

Then we can compute a posterior distribution of predictions $p(Y_{pred}|Y) = \int p(Y_{pred}|\theta)p(\theta|Y)d\theta$.

A very famous and important qauntity that depends on the model parameters, hence we can compute, is the basic reproduction number $R_0$, defined as the expected number of secondary infections produced from one infected individual in a fully susceptible population through the entire duration of the infectious period. 

By means of Bayesian inference we can build a posterior for $R_0$, that being $p(R_0|Y)$.

## Building the model

Guess what.. need to fix the data! 

```{r data_fix, echo=TRUE, warning=FALSE}
data<-read_csv("data/SIR_data.csv")
data<- data %>% select(-X1)
data<-data[,c(1,3,2)]
missing.data<-read_csv("data/dead_recovered.csv")
missing.data<-missing.data %>% select(-"confirmed")
missing.data<-missing.data %>% select(-"confirmed_daily")
missing.data<-missing.data[-1,]
missing.data<-missing.data[-c(2,3,5),]
deaths<-missing.data$deaths
recovered<-missing.data$recovered

data$`cumulative_deaths`<-deaths
data$`cumulative_recoveries`<-recovered
#pop in millions
bogotà_pop<-7.4 
antioquia_pop<-6.4
cundimarca_pop<-2.9
caldas_pop<-0.99
valle_cauca_pop<-4.4
meta_pop<-1
risaralda_pop<-0.94
santander_pop<-2.1
quindio_pop<-0.53
tolima_pop<-1.3
cauca_pop<-1.4
casanare_pop<-0.42
boyaca_pop<-1.2

tot_pop<-N<-7.4+6.4+2.9+0.99+4.4+1+0.94+2.1+0.53+1.3+1.4+0.42+1.2
N<-N*10^6 #from millions to individuals
```

We can compute the S-I-R quantities we need as follows:

  * S := Susceptible = All - Confirmed;
  
  * I := Infected = Confirmed - Recovered - Deaths;
  
  * R := Recovered or fatal = Recovered + Deaths.

```{r data_sir, echo=TRUE, warning = FALSE}
All<-as.vector(rep(N,25))
data$S<-All-data$y
data$I<-data$y-data$cumulative_deaths-data$cumulative_recoveries
data$R<-data$cumulative_deaths+data$cumulative_recoveries

SIR.data<-data[,c(2,3,6,7,8)]
head(SIR.data)
```

Let's fix things up to fix the stan model:

```{r stan_prep, echo=TRUE, warning=FALSE}
cases<-SIR.data$y
n_days<-length(cases)
t<-seq(0, n_days, by=1)
#I.C. for the ODEs
t0<-0
t<-t[-1]
i0<-1
s0<-N-i0
r0<-0
y0 = c(S=s0, I=i0, R=r0)

Sir.data<-list(n_days = n_days,
               y0 = y0,
               t0 = t0,
               ts = t,
               N = N,
               cases = cases)

#compile the model 
Sir.model<-stan_model("stan/SIR.stan")

#run the model 
fit.model<-sampling(Sir.model, data=Sir.data)
```

Let's check the correctness of the nference of the model:

```{r sir_first_check, echo=TRUE, warning=FALSE}
pars<-c("beta", "gamma", "R0", "recovery_time")
print(fit.model, pars=pars)

stan_dens(fit.model, pars=c("beta", "gamma"), separate_chains=TRUE)
```

looking at $\hat{R}$ we can see that it is $1$ for `R0` and `recovery_time`, but it is a little higher for the other two, moreover for `beta` and `gamma` the `n_eff` is really low.. this could be a symptom of the poorly mixing of the Markov chains (maybe we need to take stronger priors). 

Indeed Markov chains do not agree much. 

```{r different_priors, echo=TRUE, warning=FALSE}

#compile the model 
Sir.model2<-stan_model("stan/SIR.stan")

#run the model 
fit.model2<-sampling(Sir.model2, data=Sir.data)

pars<-c("beta", "gamma", "R0", "recovery_time")
print(fit.model2, pars=pars)

stan_dens(fit.model2, pars=c("beta", "gamma"), separate_chains=TRUE)
```


## References

[tutorial on stan for SIR model](https://mc-stan.org/users/documentation/case-studies/boarding_school_case_study.html#ref-Mihaljevic_tutorial_2016)

[how to deduce S-I-R numbers from our data](https://www.kaggle.com/lisphilar/covid-19-data-with-sir-model)

[choice of priors](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations)
  